# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
The data is in a compressed (zip) file named **activity.zip**. The data has 
three columns, steps, date, interval 

- **steps** the number of steps taken  
- **date** the date the data was recorded  
- **interval** the 5 minute interval for the data  

```{r initialData}
# read the data from the zip file
activityData <- read.csv(unz("activity.zip", "activity.csv"),  
                         header=T, stringsAsFactor=F)
# set date as date format
activityData$date <- as.Date(activityData$date, format="%Y-%m-%d")
# create day column
activityData$day <- weekdays(activityData$date)
# get list of weekdays
y <- unique(activityData$day)
# set weekday set
weekday <- y[1:5]
# set weekend set
weekend <- y[6:7]
# recode day to weekday or weekend
activityData <- within(activityData, {day[day %in% weekday] <- "Weekday" 
                      day[day %in% weekend] <- "Weekend"})
# make it a factor
activityData$day <- factor(activityData$day)
# make a copy with only the complete cases
d1 <- complete.cases(activityData)
validActivityData <- activityData[d1,]
# summarize by date and then by interval
library(plyr)
activitySumDate <- ddply(validActivityData, .(date), summarize, 
                      totalSteps = sum(steps), 
                      meanSteps = mean(steps), 
                      medianSteps = median(steps) )
activitySumInterval <- ddply(validActivityData, .(interval), 
                             summarize, 
                             meanSteps = mean(steps) )
# find the interval with the highest mean steps value
maxSteps <- activitySumInterval[activitySumInterval$meanSteps == max(activitySumInterval$meanSteps),]
```


## What is mean total number of steps taken per day?

The mean and median steps per day are listed in the table below. Notice that the
median is 0 for all days. This is because 0 is the most frequent number of steps
in the data.
```{r listMeanMedian } 
 activitySumDate[,c(1,3,4)]
```

This is a histogram of the total steps taken by day.  There are a few days that
do not have data due primarily to the NA values. 
```{r plotDay1,fig.height=4}
library(ggplot2)
ggplot(activitySumDate, aes(x=date,y=totalSteps )) + 
      geom_histogram(stat="identity", fill="blue",width=0.8) + 
      ggtitle("Total Steps by Day") + 
      ylab("Total Steps") + xlab("Date")
```

## What is the average daily activity pattern?

The most steps occurred at interval `r maxSteps[1]` and has a mean value of `r maxSteps[2]` steps.

This is a plot of the average number of steps taken by 5 minute interval across all days. 
```{r plotInterval1,fig.height=4}
library(ggplot2)
ggplot(activitySumInterval, aes(x=interval,y=meanSteps)) + 
      geom_line() +
      geom_vline(x=c(800,930,1700), colour="blue") +
      ggtitle("Average Steps by Interval\nWith vertical lines at 8AM, 9:30AM, 5PM") +
      ylab("Mean of Steps") + xlab("Interval")
```

## Imputing missing values
To recode the missing values a function was written to set the NA value to the
mean for the corresponding interval value.  
```{r recode1}
# how many NA values are in the steps variable?
naSteps <- sum(is.na(activityData$steps))
# merge the original and the summarized interval values for use in recoding NA's
copyActivityData <- merge(activityData,activitySumInterval, by="interval")
# define function to recode NA's
newStep <- function(data){
      for (i in 1:nrow(data)) {
            if (is.na(data$steps[i])) {
                  data$steps[i] <- floor(data$meanSteps[i])
            }
      }
      return(data)
}
# use function to recode NA values to mean for that interval
newActivityData <- newStep(copyActivityData)
#
library(plyr)
newActivitySumDate <- ddply(newActivityData, .(date), summarize, 
                      totalSteps = sum(steps), 
                      meanSteps = mean(steps), 
                      medianSteps = median(steps) )
activitySumIntervalDay <- ddply(newActivityData, .(interval,day), 
                                summarize, 
                                meanSteps = mean(steps))
```

There are `r naSteps` NA values in the steps variable from the original data. 

Here is a histogram of the activity data with the recoded values for the NA steps.
```{r plotRecode,fig.height=4}
# plot the data
library(ggplot2)
ggplot(newActivitySumDate, aes(x=date,y=totalSteps)) + 
      geom_histogram(stat="identity",fill="blue", width=0.8) +
      ggtitle("Total Steps by Day") +
      ylab("Total Steps") + xlab("Date")
```


The below table shows the mean and median with the recoded values. One noticable 
difference is that the days that had NA values now show up in the table with a 
median value that is not zero. Oddly enough they all have a median of 33.5. 
```{r newlistMeanMedian } 
 newActivitySumDate[,c(1,3,4)]
```

## Are there differences in activity patterns between weekdays and weekends?
A new variable has been set to **weekend** or **weekday** based on the day of the week. 
Saturday and Sunday are used for the weekend and all others for the weekday. 

The weekday plot shows a sharp spike around 8-9am with moderate activity until 
later in the day then a drop in activity. For the weekend there is a smaller spike around 
the 8-9am time followed by generally more activity through out the day. 

```{r plotFacet,fig.height=4.5}
ggplot(activitySumIntervalDay, aes(x=interval,y=meanSteps)) + 
      geom_line() + 
      facet_grid(day ~ .) +
      geom_vline(x=c(800,930,1700), colour="blue") + 
      ggtitle("Average Steps by Interval\nWith vertical lines at 8AM, 9:30AM, 5PM") +
      ylab("Mean of Steps") + xlab("Interval") +
      theme(strip.text = element_text(face="bold"),
            strip.background = element_rect(fill="green"))
```
